Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-09-07 analysis/paper/paper.Rmd
Added code for plotting raman spectra

#Stacked bar chart with counts of samples per parish and type
ggplot(Points2, aes(x= fct_rev(parish), fill = type)) +
  geom_bar() +
  theme_minimal() +
  labs(y = "Count",
       fill = "Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.title.y=element_blank(),
    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
  scale_fill_jco() +
  coord_flip() +
  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)

@@

```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region. Dark material has a higher overall absorbance, whereas white and translucent material feature lower absorbance. Brecciated quartz, which is typically described as a translucent material, falls somewhere inbetween the above groups."}
knitr::include_graphics('./analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
```

@@

## Raman
There are two distinctive features present in the Raman data. The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

@@

The resulting PCA model (fig. \@ref(fig:nir-pca1)) shows a clear trend along the first component (81.5 %) related to the above two features. The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. 
The 2nd component (6.29 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample. In contrast, along the 3rd component those samples that feature more pronounced peaks around 700 - 800 cm-1 groups up separately from the *quartz group*.

@@

<!--```{r raman-average, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
```-->
```{r raman-average, fig.width=20, fig.height=15, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}

>>>>

#Stacked bar chart with counts of samples per parish and type
ggplot(Points2, aes(x= fct_rev(parish), fill = type)) +
  geom_bar() +
  theme_minimal() +
  labs(y = "Count",
       fill = "Type") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.title.y=element_blank(),
    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
  scale_fill_jco() +
  coord_flip() +
  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)
```

@@

<!--```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region. Dark material has a higher overall absorbance, whereas white and translucent material feature lower absorbance. Brecciated quartz, which is typically described as a translucent material, falls somewhere inbetween the above groups."}
knitr::include_graphics('./analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
```-->

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

@@

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along the first component (81.5 %) related to the above two features. The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. 
The 2nd component (6.29 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample. In contrast, along the 3rd component those samples that feature more pronounced peaks around 700 - 800 cm-1 groups up separately from the *quartz group*.

@@

```{r raman-spectra, fig.width=20, fig.height=8, fig.cap="Spectra of raman measurements sampled on a dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
#Identify file with raman spectra and import
file <- '../quantifyingquartzite/analysis/data/derived_data/raman_baseline_corrected_transposed_20210826.csv'
spectra <- read.csv2(file, sep = ";", dec = ",", header = TRUE)

#Prepare function for plotting spectra
plot_multi <- function(df, x=1, y=2, cols=y,
                       xlim=range(df[,x], na.rm = T),
                       ylim=range(df[,y], na.rm = T),
                       main="", xlab="", ylab="", ...){
  # setup plot frame
  plot(NULL,
       xlim=xlim,
       ylim=ylim,
       main=main, xlab=xlab, ylab=ylab)

  # plot all your y's against your x
  pb <- sapply(seq_along(y), function(i){
    points(df[,c(x, y[i])], col=cols[i], ...)
  })
}

plot_multi(spectra, y=120, cols = rainbow(ncol(spectra)), type='l', lwd=3,
           xlab = "Wavenumber (cm-1)", ylab = "Intensity")
```

<!--```{r raman-average, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
```-->
```{r raman-pca, fig.width=20, fig.height=15, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
